Your time series is (probably) too short

### 2017-08-19 writing/manuscript_draft.Rmd
Final version of manuscript for PeerJ preprint

Running head: 
\vspace{3 mm}

@@

Author(s): \textsc{Easton R. White$^{1,2}$} 
\vspace{3 mm}

@@

$^2$Corresponding author: eawhite@ucdavis.edu, 1 480 203 7931
\vspace{3 mm}

Number of words (including 300 words for figures): X,XXX 
Number of figures:
Number of figures in supplementary material:
\vspace{3 mm}

To be submitted to: \emph{Biological Conservation}
\vspace{3 mm}

@@

Color does not need to be used for any figures in print
\vspace{3 mm}

@@

\begin{abstract}
Long-term census data are necessary to better understand ecological processes and make management decisions. However, census data are often expensive, requiring a lot of time and resources. Little work has addressed the question: when is a time series of census data long enough to address a question of interest? <!--Here, we explore two approaches to address this question: simulations and an empirical approach.--> We determine the minimum time series length required to estimate significant increases or decreases in population abundance. Importantly, we examine the ability to detect a trend with a set level of statistical power, which is often neglected in ecological time series analyses. Using a simulation approach, we demonstrate how the minimum time series length required increases with weaker trends in abundance and with higher population variability. In addition, we examine 878 populations of vertebrate species to determine the minimum time required to detect changes in their abundance. We show that 10-15 years of continuous monitoring are required in order to achieve a high level of statistical power. Similar to simulation-approaches, the minimum time required for field census data strongly depends on trend strength, population variability, and temporal autocorrelation. <!--Perhaps surprisingly, the minimum time required did not correlate well with biological explanatory variables, like body size or generation length.--> These results point to the importance of sampling populations over long periods of time. <!--We stress the need for verifying a study has enough statistical power to accurately determine long-term population trends.--> Time series less than 10-15 years are likely under-powered and potentially misleading.
\end{abstract}

@@

Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017]. Long-term data are necessary to both understand population dynamics and to assess species extinction risk (citations). Even though many time series may now be considered "long-term" (e.g. continuous plankton recorder @Giron-Nava2017), most are still short. Time series are typically short for a variety of reasons. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017].

How long of a time series is actually necessary? This question has important implications for both research and management. Scientists need to know the time series length required to address a specific question. A short time series may lead to wrong conclusions given large natural year-to-year variability (citations). Managers need to know when action is needed for a population. Therefore, they must understand when population trend over time is actually meaningful. The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary [@Gerber1999].

An important step in experimental design is to determine the number of samples required. For any particular experiment four quantities are intricately linked: significance level ($\alpha$), statistical power, effect size, and sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. A type I error is a false positive, or incorrect rejection of a true null hypothesis. For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive. The false positive rate, or significance level ($\alpha$) is often set at 0.05. A type II error ($\beta$) is failure to detect a true trend, or failure to reject a false null hypothesis. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$). The effect size is a measure of the difference between two groups. Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data are clearly non-independent.

For time series data, two general approaches to estimating sample size are appropriate. <!--Typically, calculations of sample size are given by a formula specific to standard statistical tests (e.g. t-test). However, many ecological scenarios require the use of simulations [@Bolker2008; @Johnson2015]. As a first approach,-->Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015]. Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987]. Statistical power can then be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. For example, given a time series, when is the slope from linear regression significantly different from zero? In other words, when is the time series significantly increasing or decreasing? It is then possible to determine how power changes with a variable of interest. For example, time series are simulated for different lengths of time. From these simulations, the minimum time series length required is calculated to meet certain levels of statistical significance and power [@Bolker2008].

In addition to using simulations, empirical time series can also be used. Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, we can use these measures of power to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. In addition, @Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

We use both simulations and a database of empirical time series to determine the minimum number of years required to address several questions. We estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression. First, we estimate $T_{min}$ using a simulation approach. Then we examine 878 population time series to estimate $T_{min}$. In the supplementary material, we determine $T_{min}$ for related ideas: calculating long-term growth rates, using more complicated population models, varying statistical level and power, and the use of generalized additive models.

@@

\begin{equation}
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of the increase or decrease (i.e. the rate of increase). It is important to note that any population model could be subsituted for equation \ref{sim_model}, as we do in the supplementary material (Figs. \ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).

Statistical power is then the fraction of simulations that meet some criteria. Here, we use whether the slope parameter from linear regression is significant. Statistical power changes with the simulation length (Fig. \ref{fig:theoretical_approach}b). Then, the minimum time required ($T_{min}$) is the number of years needed to obtain statistical power above some threshold (historically at 0.8, @Cohen1992). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend. We also tested the effect of varying both the significance level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015] to compare the predictability of growth rates amongst populations. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The data also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->

@@

<!--Our objective is to determine the minimum time series length ($T_{min}$) required to answer a specific question of interest. As we describe below, the minimum time series length calculation is particular to each question. Intuitively, we might expect a longer minimum time series length ($T_{min}$) required given more complicated questions (citations).-->

@@

In the supplementary material, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs. \ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig. \ref{fig:gam_example}). We found that although we obtain a similar distribution of minimum times required for GAMs, the minimum time required for GAMs is on average shorter than for linear regression (Fig. \ref{fig:min_time_dist_gam}).

@@

We determined the minimum time series length ($T_{min}$) required to address a particular question of interest. What is the minimum time series length required to determine, via linear regression, the long-term population trend? Here, the minimum time series length required had high enough statistical power ($1-\beta$ greater than 0.8) for a set significance level ($\alpha$) of 0.05. It is also possible to alter $1-\beta$ and $\alpha$. Predictably, as we increased our level of power or decreased $\alpha$, $T_{min}$ increased (Fig. \ref{fig:min_time_vs_alpha_beta}). We investigated this question through two approaches. We briefly describe results from the simulation approach and then discuss our empirical approach.

@@

We chose a simple model, but any other population model could be used (see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice should be tailored to the population of interest. We explored how the simulation approach can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}). Similar to the simulation approach decribed above, the minimum time required for the lemon shark population was strongly dependent on model parameters.

@@

```{r,echo=F,fig.cap='Distributions of the minimum time required for four different biological classes. The minimum time required calculation corresponds to a significant level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (interannual variability in population size), and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}). Estimates of $T_{min}$ varied between biological class (Fig. \ref{fig:min_time_dist}). Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years. Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling. Differences between these classes were explained by differences in interannual variability in population size (Fig. \ref{fig:class}). We also examined a subset of populations where life-history characteristics were known. None of these explanatory variables were predictive of the minimum time series length required (Fig. \ref{fig:biological_correlates}).

The minimum time series length required was strongly correlated with trend strength, population variance, and autocorrelation (Fig. \ref{fig:correlates}). This is in line based on our simulations and those of others [@Rhodes2011]. Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table A1). Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)`% of the explained deviance, which is similar to $R^2$ [@Zuur2009], in minimum time series length required.

@@

We explored two different approaches to estimate the minimum time series length required to address a particular question of interest. We specifically asked, what is the minimum time series length required to determine long-term population trends using linear regression? This is one of the simplest questions one could ask of a time series. We would expect more involved questions (e.g. detecting chaos) to require a larger minimum time series length. The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987].

Here, we focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time. We examined long time series and then subsampled each to determine the minimum time required to achieve a desired level of significance and power for linear regression. Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006]. We found that for `r nrow(pop_info)` populations, at least 10-15 years of continuous monitoring were necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig. \ref{fig:biological_correlates}). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have been a result of at least two effects. First, the database we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time. Therefore, life-history characteristics may be more important for other questions (e.g. species extinction risk) where estimates minimum time required are needed.

There has been interest in determining the minimum time series length in the bird research community [@Hatch2003; @Bart2004; @Seavy2007]. For example, @Hatch2003 used seabird monitoring data to estimate the minimum time needed to achieve appropriate levels of power. He found that the minimum time required ranged between 11 and 69 years depending on species, trend strength, and study design. Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations. They found that only about 42% of populations in the Breeding Bird Survey met their criteria. @Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision. For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.

@@

@@

Our work has some limitations in determining the minimum time series length required. First, $T_{min}$ is particular to the specific question of interest. For example, in the supplementary material, we used generalized additive models to detect population trends and came up with a similar, but quantitatively different value for $T_{min}$ (Fig. \ref{fig:min_time_dist_gam}). An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig. \ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a specific population model parameterized for each population of interest. Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest. Clearly, this is not always practical, especially if conducting analyses for a wide array of species as we do here. In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression (Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to determine population trends. This goes beyond previous work that either focused on theoretical investigations or a limited number of species. We show that to identify long-term changes in abundance, 10-15 years of continuous monitoring are required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions [@Gerrodette1987], we also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e. rate of increase), variability in population size, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, census data less than 10-15 years are probably not reliable. This stresses the importance of long-term monitoring programs. From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short we lack statistical power to reliably detect long-term population trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work should examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.

@@

In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends. All code and data can be found at www.github.com/erwhite1

>>>>

@@

Author: \textsc{Easton R. White$^{1,2}$} 
\vspace{3 mm}

@@

$^2$Corresponding author: eawhite@ucdavis.edu <!--, 1 480 203 7931-->
\vspace{3 mm}

Number of words (including 300 words per figure):  5900 
\vspace{1 mm}

Number of figures: 3 
\vspace{1 mm}

Supplementary material: 11 figures and 1 table 
\vspace{3 mm}

Data and code for all the figures and tables can be found at (https://github.com/erwhite1/time-series-project).

<!--To be submitted to: \emph{Biological Conservation}
\vspace{3 mm}

@@

Color does not need to be used for any figures in print
\vspace{3 mm}-->

@@

\begin{abstract}
Long-term time series are necessary to better understand population dynamics, assess species' conservation status, and make management decisions. However, population data are often expensive, requiring a lot of time and resources. When is a population time series long enough to address a question of interest? We determine the minimum time series length required to detect significant increases or decreases in population abundance. To address this question, we use simulation methods and examine 878 populations of vertebrate species. Here we show that 15-20 years of continuous monitoring are required in order to achieve a high level of statistical power. For both simulations and the time series data, the minimum time required depends on trend strength, population variability, and temporal autocorrelation. These results point to the importance of sampling populations over long periods of time. We argue that statistical power needs to be considered in monitoring program design and evaluation. Time series less than 15-20 years are likely underpowered and potentially misleading.
\end{abstract}

@@

Observational studies and population time series have become a cornerstone of modern ecological research and conservation biology [@Magurran2010; @Hughes2017]. Long-term data are necessary to both understand population dynamics and to assess species extinction risk. Even though many time series may now be considered "long-term" (e.g. continuous plankton recorder, @Giron-Nava2017), most are still short. Time series are typically short for a variety of reasons [@Field2007]. They are often coupled with an experiment, which may only last a couple of years. In addition, short funding cycles make it difficult to examine populations over longer periods of time [@Hughes2017].

How long of a time series is actually necessary? This question has important implications for both research and management [@Nichols2006]. Scientists need to know the time series length required to address a specific question. A short time series may lead to wrong conclusions given large natural year-to-year variability. Managers need to know when action is needed for a population. Therefore, they must understand when population trend over time is actually meaningful. The IUCN Red List Categories and Criteria suggest, under Criterion A2, a species qualifies as vulnerable if it has experienced a 30% decline over 10 years, or 3 generations [@IUCN2012]. For both scientific and management questions, because sampling is typically expensive, we also do not want to sample for longer than is necessary. For example, @Gerber1999 investigated the minimum time series required to estimate population growth of the endangered, but recovering, eastern North Pacific gray whale (\emph{Eschrichtius robustus}). They used a long-term census to retroactively determine the minimum time series required to assess threat status. They found that only 11 years were needed, eight years before the delisting decision was made. This highlights the importance of estimating the minimum time series required as an earlier decision would have saved time and money [@Gerber1999].

An important step in experimental design is to determine the number of samples required. For any particular experiment four quantities are intricately linked: significance level ($\alpha$), statistical power, effect size, and sample size [@Legg2006]. The exact relationship between these quantities depends on the specific statistical test. A type I error is a false positive, or incorrect rejection of a true null hypothesis. For example, if a time series was assessed as significantly increasing or decreasing---when there was no true significant trend---this would be a false positive. The false positive rate, or significance level ($\alpha$) is often set at 0.05 (although this is purely historical, @Mapstone1995). A type II error ($\beta$) is a failure to detect a true trend, or failure to reject a false null hypothesis. Formally, statistical power ($1-\beta$) is one minus the probability of a type II error ($\beta$). The effect size is a measure of the difference between two groups. Prior to an experiment, one could set appropriate levels of power, significance level, and the effect size to estimate the sample size required for the experiment. This approach, however, is not straight-forward for a time series, or more complicated scenarios [@Johnson2015], as data are clearly non-independent.

For time series data, two general approaches to estimating sample size are appropriate. <!--Typically, calculations of sample size are given by a formula specific to standard statistical tests (e.g. t-test). However, many ecological scenarios require the use of simulations [@Bolker2008; @Johnson2015]. As a first approach,-->Simulations can be designed for a specific population and question [@Bolker2008; @Johnson2015]. Simple models can be simulated with parameter values corresponding to a population of interest [@Gerrodette1987]. Statistical power can then be calculated as the fraction of simulations that meet some criteria. The specific criteria depend on the question at hand. For example, given a time series, when is the slope from linear regression significantly different from zero? In other words, when is the time series significantly increasing or decreasing? It is then possible to determine how power changes with a variable of interest. For example, time series can be simulated for different lengths of time. From these simulations, the minimum time series length required is calculated to meet certain levels of statistical significance and power [@Bolker2008].

In addition to using simulations, empirical time series can also be used. Multiple replicates of similar populations are usually not available, but it is possible to subsample an empirical time series [@Gerber1999]. Subsamples of different lengths can then be evaluated to see which fraction of subsamples meet some criteria, again a measure of statistical power. Similar to the simulation approach, these measures of power can be used to determine the minimum time series required for a particular question of interest.

Past work has investigated questions related to the minimum time series required to estimate trends in population size over time [@Wagner2009; @Giron-Nava2017]. For example, @Rhodes2011 examined the optimal allocation of effort between spatial and temporal replicates. Using simple populations models, they found that the allocation of effort depends on environmental variation, spatial and temporal autocorrelation, and observer error. @Rueda-Cediel2015 also used a modeling approach, but parameterized a model specifically for a threatened snail, \emph{Tasmaphena lamproides}. They found that for this short-lived organism, 15 years was adequate to assess long-term trends in abundance. @Hatch2003 examined seabird populations and determined the minimum time required to achieve high statistical power ranged between 11 and 69 years. However, these studies, and other past work, have typically focused either on theoretical aspects of monitoring design or focused on only a few species.

We use both simulations and empirical time series to determine the minimum number of years required to address several questions. We estimate the minimum time series length required ($T_{min}$) to assess long-term changes in abundance via simple linear regression. First, we estimate $T_{min}$ using a simulation approach. Then we examine 878 population time series to estimate $T_{min}$. In the supplementary material, we determine $T_{min}$ for related ideas: calculating long-term growth rates, using more complicated population models, varying statistical level and power, and the use of generalized additive models.

@@

\begin{equation}
N(t + 1) = N(t) + r(t) + \epsilon \mbox{ with } \epsilon \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of the increase or decrease (i.e. the rate of increase). It is important to note that any population model could be substituted for equation \ref{sim_model}, as we do in the supplementary material (Figs. \ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).

Statistical power is then the fraction of simulations that meet some criteria. Here, our criteria is whether the slope parameter from linear regression is significant at the 0.05 threshold. Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend. We also tested the effect of varying both the significance level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]; they compared the predictability of growth rates among populations. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The data also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. All analyses were conducted in R [@RCoreTeam2016]. <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->

@@

@@

In the supplementary material, we show how the same approach described here can be used to determine the minimum time required to estimate a population's geometric growth rate (Figs. \ref{fig:growth_rate},\ref{fig:min_time_growth_dist}). We also determine the minimum time required to estimate long-term trends according to generalized additive models, instead of the simple linear models used here (Fig. \ref{fig:gam_example}).

@@

We determined the minimum time series length ($T_{min}$) required to address a particular question of interest. What is the minimum time series length required to determine, via linear regression, the long-term population trend? Here, the minimum time series length required had high enough statistical power ($1-\beta$ greater than 0.8) for a set significance level ($\alpha$) of 0.05. It is also possible to alter $1-\beta$ and $\alpha$. Predictably, as we increased our level of power or decreased $\alpha$, $T_{min}$ increased (Fig. \ref{fig:min_time_vs_alpha_beta}). We estimated $T_{min}$ using two approaches. We briefly describe results from the simulation approach and then discuss our empirical approach.

@@

We chose a simple model, but any other population model could be used (see example in Fig. \ref{fig:min_time_dist_log_pop}). Ideally, the specific model choice should be tailored to the population of interest. We explored how the simulation approach can be applied to more biologically-realistic population models (Fig. \ref{fig:shark_example}). More specifically, we determined the minimum time required to estimate long-term population trends using a stochastic, age-structured model of lemon shark population dynamics in the Bahamas [@White2014]. We found that over 27 years of continuous monitoring were needed in this particular scenario (Fig. \ref{fig:shark_example}). Similar to the simulation approach described above, the minimum time required for the lemon shark population was strongly dependent on model parameters.

@@

```{r,echo=F,fig.cap='Distributions of the minimum time required for four different biological classes. The minimum time required calculation corresponds to a significance level of 0.05 and statistical power of 0.8.\\label{fig:min_time_dist}',fig.width=6}
# Pull in analysis of average min time needed for **empirical species** (save analysis as Rdata file or something)
 load("../cleaned-data/cleaned_timeseries_database_with_min_time_linear_regression.Rdata")

@@

```{r,echo=F, fig.cap='Minimum time required to estimate change in abundance correlated with (a) trend strength (absolute value of slope coefficient estimated from linear regression), (b) population variance (inter-annual variability in population size), and (c) temporal lag-1 autocorrelation.\\label{fig:correlates}',fig.height=7, fig.width=4}

@@

Across all the populations we examined, we found an average minimum time series length required ($T_{min}$) of `r round(mean(pop_info$min_time_for_power),1)` ($\sigma$=`r round((var(pop_info$min_time_for_power))^0.5,1)`), with a wide distribution (Fig. \ref{fig:min_time_dist}). Estimates of $T_{min}$ varied between biological class (Fig. \ref{fig:min_time_dist}). Ray-finned fish (class Actinopterygii) typically had estimates of $T_{min}$ over 20 years. Birds (class Aves) had a much wider distribution of $T_{min}$, but usually required less years of sampling. Differences between these classes were explained by differences in inter-annual variability in population size (Fig. \ref{fig:class}). We also examined a subset of populations where life-history characteristics were known. None of these explanatory variables were predictive of the minimum time series length required (Fig. \ref{fig:biological_correlates}).

The minimum time series length required was strongly correlated with trend strength, population variance, and autocorrelation (Fig. \ref{fig:correlates}). This is in line based on our simulations and those of others [@Rhodes2011]. Using a generalized linear model, with a Poisson error structure, all three of these explanatory variables were significant and had large effect sizes (see Table A1). Combined, trend strength, population variance, and autocorrelation account for about `r round(100*(poisson_model$null.deviance - poisson_model$deviance)/(poisson_model$null.deviance),1)`% of the explained deviance [@Zuur2009] in minimum time series length required.

@@

We explored two approaches to estimate the minimum time series length required to address a particular question of interest. We asked, what is the minimum time series length required to determine long-term population trends using linear regression? This is one of the simplest questions one could ask of a time series. The simulation-based approach has been suggested by others, especially in situations more complicated than that suited for classic power analysis [@Gerrodette1987;@Johnson2015;@Bolker2008]. Our simulations support past work that longer time series are needed when the trend strength (i.e. rate of increase or decrease) is weak or when population variability is high [@Gerrodette1987]. We also showed how the simulation model can be altered for a particular population (Fig. \ref{fig:shark_example}) or question (Figs. \ref{fig:min_time_dist_log_pop},\ref{fig:gam_example}).

Here, we focus on an empirical approach to estimate the minimum time series length required to assess changes in abundance over time. We examined `r nrow(pop_info)` population time series (all longer than 35 years). We then subsampled each to determine the minimum time required to achieve a desired significance level and power for linear regression. Statistical power is important as it provides on information as to the necessary samples required to determine a significant trend [@Legg2006]. We found that at least 15-20 years of continuous monitoring were typically necessary (Fig. \ref{fig:min_time_dist}). This time frame is in line with past work on a short-lived snail species [@Rueda-Cediel2015] and a long-lived whale species [@Gerber1999]. @Hatch2003 used seabird monitoring data to estimate minimum sampling requirements. He found that the time required ranged from 11 to 69 years depending on species, trend strength, and study design.

In line with theoretical predictions [@Rhodes2011], we also found $T_{min}$ was strongly correlated with the trend strength, variability in population size, and temporal autocorrelation (Fig. \ref{fig:correlates}). Contrary to our prior expectations, we also found that $T_{min}$ did not correlate with any biological variables of interest (Fig. \ref{fig:biological_correlates}). We initially hypothesized that species with longer lifespans or generation times may require a longer sampling period. Our result could have been a result of at least two factors. First, the data we used may not include a diverse enough set of species with different life history traits. Second, the question we poised, whether a population is increasing or decreasing, was specifically concerned with trends in population density over time. Therefore, life-history characteristics may be more important for other questions, like estimating species extinction risk [@Hutchings2012].

<!-- Similarly, @Bart2004 recommended a standard of 80% power to detect a 50% decline over 20 years in landbird populations. They found that only about 42% of populations in the Breeding Bird Survey met their criteria. @Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision. For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs.-->

@@

@Seavy2007 asked whether statistical power was even a useful framework for assessing long-term population trends. They used 24 years of census data on Red-tailed Tropicbirds (\emph{Phaethon rubricauda}) in Hawaii and showed that to detect a 50% decline over 10 years almost always resulting in high statistical power (above 0.8). Therefore, they cautioned against only using power analyses to design monitoring schemes and instead argued for metrics that would increase precision. For example, @Seavy2007 suggest improving randomization, reducing bias, and increases detection probability when designing and evaluating monitoring programs. We agree that power analyses should not be the only consideration when designing monitoring schemes. However, unlike @Seavy2007, our results indicate that longer than 10 years is often needed to achieve high statistical power. Therefore, the design of monitoring programs should include calculations of statistical power, the allocation of sampling in space versus time [@Rhodes2011], and metrics to increase precision. Ideally, a formal decision analysis to evaluate these different factors would be conducted to design or assess any monitoring program [@Hauser2006;@McDonald-Madden2010].

@@

Our work has some limitations in determining the minimum time series length required. First, $T_{min}$ is particular to the specific question of interest. An additional complication is that for our empirical approach, the subsampling of the full time series allows for estimates of power, but the individual subsamples are clearly not independent of one another. Further, estimates of $T_{min}$ depend on chosen values of $\alpha$ and $\beta$ (Fig. \ref{fig:min_time_vs_alpha_beta}). In an ideal setting, we would build a specific population model parameterized for each population of interest. Then, model simulations could be used to estimate the minimum time series required to address each specific question of interest. Clearly, this is not always practical, especially if conducting analyses for a wide array of species as we do here. In addition, our statistical models suggest that $T_{min}$ does not correlate with any biological variables of interest, at least for the question of linear regression (Fig. \ref{fig:biological_correlates}). Therefore, it is not possible to use these results to predict $T_{min}$ for another population, even if the population is of a species with a similar life-history to one in our database.

@@

We use a database of `r nrow(pop_info)` populations to determine the minimum time series length required to detect population trends. This goes beyond previous work that either focused on theoretical investigations or a limited number of species. We show that to identify long-term changes in abundance, 15-20 years of continuous monitoring are often required (Fig. \ref{fig:min_time_dist}). In line with theoretical predictions [@Gerrodette1987], we also show that $T_{min}$ is strongly correlated with the long-term population trend (i.e. rate of increase), variability in population size, and the temporal autocorrelation (Fig. \ref{fig:correlates}). Our work implies that for many populations, time series less than 15-20 years are probably not reliable for detecting population trends. This result highlights the importance of long-term monitoring programs. From both a scientific and management perspective estimates of $T_{min}$ are important. If a time series is too short, we lack statistical power to reliably detect long-term population trends. In addition, a time series that is too long may be a poor use of already limited funds [@Gerber1999]. Future work should examine other species, with a wider range of life history characteristics. In addition, similar approaches can be used to determine the minimum time series length required to address additional questions of interest.

@@

In the supporting material, we provide an expanded methods sections, additional figures, minimum time calculations for determining exponential growth, simulations with a more complicated population model, minimum time calculations for determining long-term growth rates, and the use of generalized additive models to identify population trends. All code and data can be found at https://github.com/erwhite1/time-series-project
